---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

First, we load the data and look at the data structure and formats.
```{r}
dane0<-read.csv('activity.csv')
str(dane0)
```

We change the format of *date* from factor to date.
```{r}
dane0$date<-as.Date(as.character(dane0$date))
str(dane0)
```

## What is mean total number of steps taken per day?
First, we aggregate the data using functions from dplyr package.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
stepsPerDay<-dane0 %>% group_by(date) %>% summarise(steps_count=sum(steps))
```

In order to make histogram, we use ggplot2 package.
```{r, warning=FALSE, message=FALSE, fig.height=4, fig.align='center'}
library(ggplot2)
g<-ggplot(stepsPerDay, aes(x=steps_count))+geom_histogram()
plot(g)
```

To report mean and median, we may use summary function.
```{r}
summary(stepsPerDay$steps_count)
```

## What is the average daily activity pattern?
First, we aggregate the date using dplyr and then make a plot with base graphics.
```{r, warning=FALSE, fig.height=4, fig.align='center'}
stepsPerInterval<-dane0 %>% group_by(interval) %>% summarise(avg_steps=mean(steps, na.rm=TRUE))
interval_max_steps<-stepsPerInterval[which.max(stepsPerInterval$avg_steps),1]
interval_max_steps
plot(x=stepsPerInterval$interval, y=stepsPerInterval$avg_steps, type="l")
abline(v=stepsPerInterval[which.max(stepsPerInterval$avg_steps),1], type="l", col="red")
```
The interval with highest average number of steps is 835- this is indicated on the plot with red vertical line.

## Imputing missing values
First, it is worth checking how many missing values we have in our dataset
```{r}
sapply(dane0, function(x) sum(is.na(x)))
sapply(dane0, function(x) sum(is.na(x))/nrow(dane0))
```
There are about 13% of NA values in *steps* variable. As the average number of steps varies significantly between intervals, we will impute missing values with average number of steps per interval.

```{r}
dane1<-merge(dane0, stepsPerInterval, by="interval")
dane1$steps<-ifelse(is.na(dane1$steps), dane1$avg_steps, dane1$steps)
```

We compare the impact of imputation looking at the statistics for steps count per day.

```{r}
stepsPerDay1<-dane1 %>% group_by(date) %>% summarise(steps_count=sum(steps))
summary(stepsPerDay1$steps_count)
```
The quantiles slightly differ as we have imputed values for 8 days that were initially missing. As we have filled in missing values with average values for each interval, the global mean has not changed- in each of the 8 days that were missing there are average values observed for the remaining days. The histogram of the total number of steps taken each day is presented below.

```{r, warning=FALSE, message=FALSE, fig.height=4, fig.align='center'}
g2<-ggplot(stepsPerDay1, aes(x=steps_count))+geom_histogram()
plot(g2)
```

## Are there differences in activity patterns between weekdays and weekends?
First, we create new factor variable indicating whether a given date is a weekday or weekend day.

```{r}
dane1$day<-weekdays(dane1$date)
dane1$day<-ifelse(dane1$day %in% c("sobota", "niedziela"), "weekend", "weekday")
dane1$day<-as.factor(dane1$day)
```

In order to make a plot, we first aggregate data using dplyr and then utilise ggplot2.

```{r, fig.align='center', fig.height=4}
stepsPerWeekday<- dane1 %>% group_by(interval,day) %>% summarise(avg_steps=mean(steps))
g3<-ggplot(stepsPerWeekday, aes(x=interval, y=avg_steps, col=day))+geom_line(lwd=1)+facet_grid(day~.)
plot(g3)
```
It seems that activity begins later on weekend than on weekdays. On weekdays there is a substantial peak around interval 835, while on weekend activity is ditributed more uniformly.





